Parsing Postfix log files

### 2009-04-04 doc/logparser-implementation.tex
Corrections/improvements based on Carl's comments. Improved the flow chart, rewriting labels and making it easier to read.

This chapter documents the implementation of \parsername{}, based on the
architecture described in \sectionref{parser architecture}.  Any design may
look good when examined in the abstract, but the real test of the design
comes with the first concrete implementation; only then do any practical
difficulties come to light.  Implementing this architecture was relatively
straightforward, with only minor changes required: rule conditions and
cascaded parsing were added, although later the implementation stopped
using cascaded parsing.  The difficulties came not from the architecture
and design, but from anomalies in the log files and Postfix's behaviour, as
described in \sectionref{complications}.

\parsername{} deals with all the complications of parsing, the
eccentricities and oddities of Postfix log files, presenting the resulting
data in a normalised, simple to use representation.  The parser's task is
to follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.

@@

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users; that
section finishes with the features that are important for this thesis but
not strictly necessary for parsing: the performance data collected by the
framework, the optimisations that can be disabled to show their effect, and
the debugging options it provides.  The second component of the
architecture is the actions: their implementation is documented, with a
graph of how often each action is specified by rules, and a description of
why some actions are more popular than others.  Every action that is part
of \parsername{} is documented, and the actions section concludes with the
process of adding a new action, and how to determine the values for each of
the required fields.  Rules are the final component of the architecture,
and also the most visible to users, because it is likely that users will
need to add their own rules to recognise their log lines.  An example rule
used by \parsername{} is shown, with every field clearly documented; this
is followed by a description of adding new rules, and how to determine the
value of each of the fields.  \parsername{} provides a utility to create
regexes from unrecognised log lines; the algorithm it uses is documented,
including the differences between it and the original algorithm.  The rules
section finishes with a discussion of how \parsername{} uses rule
conditions and overlapping rules, plus a description of the regex snippets
provided to ease the process of writing regexes.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering from a deluge of spam
        or a mail loop.

@@

A picture is said to be worth a thousand words: the flow chart in
\figureref{flow chart image} shows the most common paths a connection or
mail can take through \parsername{}; the difficulties described in
\sectionref{complications} are excluded for the sake of clarity.  The flow
chart is intended to be a graphical overview of how a mail progresses
through both Postfix and \parsername{}, providing an overall context into
which the detailed descriptions in the remainder of this chapter will fit,
in particular the actions (\sectionref{actions in implementation}) and
complications (\sectionref{complications}).

@@

The intermingling of log entries from different mails immediately rules out
the possibility of handling each mail in isolation; the parser must be able
to handle multiple mails in parallel, each potentially at a different stage
in its journey, without any interference between mails --- except in the
minority of cases where intra-mail interference is required.  The best way
to implement this is to maintain state information for every unfinished
mail and manipulate the appropriate mail correctly for each log line
encountered.

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or generated internally by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
the mail moves on to the delivery stage.  Re-injection due to forwarding
sadly lacks explicit log lines of its own; it is explained fully in
\sectionref{Re-injected mails}.  Internally generated mails lack any
explicit origin in Postfix 2.2.x and must be detected using heuristics as
described in \sectionref{identifying bounce notifications}; later versions
of Postfix do provide log lines for internally generated mails.  Bounce
notifications are the primary example of internally generated mails, though
there are other types, e.g.\ Postfix may generate mails to the
administrator when it encounters configuration errors, but such mails are
presumably rare.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, etc.\
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; or some interleaving of both.

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the resulting data will be
        saved to the database; if there were no rejections there will not
        be any data to save.  Any mails accepted will already have been
        cloned so their data is in another data structure, nd will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked by the parent mail \flowchart{TRACK}{9}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn, e.g.\
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere, analogous to the
        overall similarity and minor differences amongst Irish English,
        British English, American English, and Australian English.
        Depending on the database in use there may be tools available that
        reduce or remove the requirement to know \acronym{SQL}.

@@

    \item [warning] Postfix can be configured to log a warning instead of
        enforcing a restriction that would reject an \acronym{SMTP} command
        --- a mechanism that is quite useful for testing new restrictions.
        This field will be false for a real rejection, or true if the log
        line was a warning.  This field should be ignored if the result is
        not a rejection, i.e.\ the action field of the associated rule is
        not \action{DELIVERY\_REJECTED}.

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(see \sectionref{rule conditions in implementation}), falling back to
generic rules if necessary, and warning if the log line is unrecognised.
The rules are sorted for increased efficiency; see \sectionref{rule
ordering for efficiency} for details.  The repetitive nature of log files
gives them high compression ratios; the framework supports reading
compressed log files, simplifying the process of parsing them because they
do not have to be uncompressed to a temporary file before parsing begins.
The framework also displays a progress bar to show how far parsing has
progressed through the log file, and to give the user an indication of how
long the remainder of the parsing process is likely to take.  The progress
bar is not as reliable when parsing compressed log files, because a
compressed block may uncompress to a variable number of log lines; there is
also variation between the recognition and processing time of individual
log lines, but overall the progress bar is a useful addition.

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \action{SAVE\_DATA}, the action
that saves useful information without doing any other processing.  The
third most common action is, perhaps surprisingly, \action{UNINTERESTING}:
this action does nothing when invoked, allowing uninteresting log lines to
be parsed without any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\action{UNINTERESTING} action recognise log lines that are not associated
with a specific mail, e.g.\ notices about configuration files changing;
these log lines are recognised and processed so that the framework can warn
about unrecognised log lines, informing the user that they need to add to
or extend their ruleset.  Most of the remaining actions have only one or
two associated rules: some are specialised actions required to address a
deficiency in the log files, or a complication that arises during parsing;
other actions will only ever have one log line variant, e.g.\ all log lines
showing that a remote client has connected are recognised by a single rule
and handled by the \action{CONNECT} action.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client.  This is the action
        most frequently specified by rules, because so many different
        restrictions are used to reject delivery attempts.  This action is
        quite simple: if the log line contains a queueid, save the data
        captured by the rule's regex to the mail identified by that
        queueid; otherwise save it to the connection identified by the
        \acronym{pid} in the log line.  No further processing of the log
        line is required.

@@

    \item [MAIL\_DISCARDED] Sometimes mail is discarded, e.g.\ mail
        submitted locally that is larger than the limit configured by the
        administrator.  This action is used for those mails; it invokes the
        \action{COMMIT} action, but is a separate action to simplify
        further analysis.

@@

    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies, is killed by a
        signal, or exits unsuccessfully; the associated connection must be
        cleaned up, entered in the database if it has sufficient data, and
        deleted from the state tables.  In some cases the connection will
        not have enough data to satisfy the database schema, so it cannot
        be entered into the database; unfortunately this means that some
        data is lost.

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their new action.  The
action may need to extend the list of valid combinations described in
\sectionref{out of order log lines} if the new action creates a different
set of acceptable programs, but this would only be necessary if the new
action processes log lines from Postfix components \parsername{} currently
does not have rules for, e.g.\ \daemon{virtual} or \daemon{lmtp}.

@@

There are other attributes that may be required to make a rule work
correctly: connection\_data, result\_data, priority, or restriction\_name.
In general it will only become clear that connection\_data or result\_data
are required when \parsername{} warns about an entry in the connections or
results tables that is missing some required fields, because values for the
missing fields are not present in any of the log lines.  E.g.\ the rule
recognising the \daemon{pickup} component processing a mail sets the
client\_hostname to \texttt{localhost} and the client\_ip to
\texttt{127.0.0.1}, because the mail originates on the local server.  If
the new rule deliberately uses the architecture's overlapping rules feature
it will be obvious that the priority field needs to be set, on this rule
and possibly others; the priority field may be needed on unintentionally
overlapping rules too, but that is more difficult to determine.  Finally
the restriction\_name field should be set when the rule's action is
\action{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
clear from the content of the log line.

@@

Rule conditions as part of the architecture have already been documented in
\sectionref{rule conditions in architecture}, and they can be very complex
and difficult to evaluate.  In contrast, \parsername{} uses very simple
rule conditions: each rule has a \texttt{program} attribute that specifies
the Postfix component whose log lines it recognises, and each log line
contains the name of the Postfix component that produced it; when trying to
recognise log lines, the framework will only use rules where the two are
equal.  This avoids needlessly trying rules that will not recognise the log
line, or worse, might recognise it unintentionally.  In addition the
framework supports generic rules, whose program attribute is ``*''; these
will be used if none of the program-specific rules recognise the log line.
If none of the rules are successful the framework will warn the user.

@@

>>>>

This chapter documents the implementation of \parsername{}, based on the
architecture described in \sectionref{parser architecture}.  Any design may
look good when examined in the abstract, but the real test of the design
comes with the first concrete implementation; only then do any practical
difficulties come to light.  Implementing this architecture was relatively
straightforward: the difficulties came not from the architecture and
design, but from anomalies in the log files and Postfix's behaviour, as
described in \sectionref{complications}.  \parsername{} successfully deals
with all the difficulties that were discovered during its development, but
further difficulties may arise during future usage; unfortunately solving
all potential difficulties that may arise is an intractable task, but the
descriptions of the solutions developed thus far (see
\sectionref{complications}) should help if someone needs to develop a
solution of their own.

\parsername{} deals with all the complications of parsing, the
eccentricities and oddities of Postfix log files, presenting the resulting
data in a normalised, simple to use representation.  The parser's task is
to follow the journey each mail takes through Postfix, combining the data
captured by rules into a coherent whole, saving it in a useful and
consistent form, and performing housekeeping duties.  The intermingling of
log entries from different mails immediately rules out the possibility of
handling each mail in isolation; the parser must be able to handle multiple
mails in parallel, each potentially at a different stage in its journey,
without any interference between mails --- except in the minority of cases
where intra-mail interference is required, e.g.\ mail re-injected for
forwarding (\sectionref{Re-injected mails}).  The best way to deal with
intermingling is to maintain state information for every unfinished mail
and manipulate the appropriate mail correctly for each log line
encountered.

@@

The next three sections document the implementation of the three components
of the architecture.  First is the framework, including its initialisation
phase, the parsing process, and the conveniences it offers to users; that
section finishes with the features that are important for this thesis but
not strictly necessary for parsing: the performance data collected by the
framework, the optimisations that can be disabled to show their effect, and
the debugging options it provides.  The second component of the
architecture is the actions: their implementation is documented, with a
graph of how often each action is specified by rules, and a description of
why some actions are more popular than others.  Every action that is part
of \parsername{} is documented, and the actions section concludes with the
process of adding a new action, and how to determine the values for each of
the required fields.  Rules are the final component of the architecture,
and also the most visible to advanced users, e.g.\ systems administrators,
because it is likely that they will need to add their own rules to
recognise their log lines.  An example rule used by \parsername{} is shown,
with every field clearly documented; this is followed by a description of
adding new rules, and how to determine the value of each of the fields.
\parsername{} provides a utility to create regexes from unrecognised log
lines; the algorithm it uses is documented, including the differences
between it and the original algorithm.  The rules section finishes with a
discussion of how \parsername{} uses rule conditions and overlapping rules,
plus a description of the regex snippets provided to ease the process of
writing regexes.

@@

    \item The log files are whole and complete: nothing has been removed,
        either deliberately or accidentally (e.g.\ log file rotation gone
        awry, file system filling up, logging system unable to cope with
        the volume of log messages).  On a well run system it is extremely
        unlikely that any of these problems will arise, though it is of
        course possible, particularly when suffering from a deluge of spam
        or a mail loop.  When parsing individual log files in isolation,
        it is highly likely that some mails will have log lines in previous
        log files, and others will have log lines in subsequent log files;
        to alleviate this problem \parsername{} supports saving and loading
        its state tables, so they can be preserved between log files.

@@

The flow chart in \figureref{flow chart image} shows the most common paths
a connection or mail can take through \parsername{}; the difficulties
described in \sectionref{complications} are excluded for the sake of
clarity.  The flow chart is intended to be a graphical overview of how a
mail progresses through both Postfix and \parsername{}, providing an
overall context into which the detailed descriptions in the remainder of
this chapter will fit, in particular the actions (\sectionref{actions in
implementation}) and complications (\sectionref{complications}).  The
states and transitions shown in \figureref{flow chart image} will be
explained later in this section.

@@

Everything starts off with a mail entering the system, whether by local
submission via \daemon{postdrop}, by \acronym{SMTP}, by re-injection due to
forwarding, or generated internally by Postfix.  Local submission is the
simplest case: a queueid is assigned immediately \flowchart{PICKUP}{2}, and
the mail moves on to the delivery stage.  Re-injection due to forwarding
lacks explicit log lines of its own; it is explained fully in
\sectionref{Re-injected mails}.  Internally generated mails lack any
explicit origin in Postfix 2.2.x and must be detected using heuristics as
described in \sectionref{identifying bounce notifications}; later versions
of Postfix do provide log lines for internally generated mails
\flowchart{BOUNCE\_CREATED}{3}.  Bounce notifications are the primary
example of internally generated mails, though there are other types, e.g.\
Postfix may generate mails to the administrator when it encounters
configuration errors.

@@

    \item This is followed by rejection of sender address, recipient
        addresses, client \acronym{IP} address or hostname, etc.\
        \flowchart{DELIVERY\_REJECTED}{4}; acceptance of one or more mails
        \flowchart{CLONE}{5}; or some random interleaving of both.

    \item The client disconnects \flowchart{DISCONNECT}{6}.  If Postfix has
        rejected any \acronym{SMTP} commands the resulting data will be
        saved to the database; if there were no rejections there will not
        be any data to save.  Any mails accepted will already have been
        cloned so their data is in another data structure, and will be
        delivered in the same way as mails that entered the system by any
        other route.

@@

    \item Sometimes mail is re-injected and the child mail needs to be
        tracked with the parent mail \flowchart{TRACK}{9}; the handling of
        re-injected mails is described in \sectionref{tracking re-injected
        mail}.

@@

    \item \acronym{SQL} is reasonably standard and many people will already
        be familiar with it; for those unfamiliar with it there are lots of
        readily available resources from which to learn, e.g.\
        \urlLastChecked{http://philip.greenspun.com/sql/}{2009/02/23}.
        Although every vendor implements a different dialect of
        \acronym{SQL}, the basics are the same everywhere.  Depending on
        the database in use there may be tools available that reduce or
        remove the requirement to know \acronym{SQL}.

@@

    \item [warning] Administrators can configure Postfix to log a warning
        instead of enforcing a restriction that would reject an
        \acronym{SMTP} command --- a mechanism that is quite useful for
        testing new restrictions.  This field will be false for a real
        rejection, or true if the log line was a warning.  This field
        should be ignored if the result is not a rejection, i.e.\ the
        action field of the associated rule is not
        \action{DELIVERY\_REJECTED}.

@@

The framework is now ready to begin parsing.  For each log line it will use
those rules whose \texttt{program} field equals the program in the log line
(see \sectionref{rule conditions in implementation}), falling back to
generic rules if necessary, and warning if the log line is unrecognised.
The rules are sorted for increased efficiency; see \sectionref{rule
ordering for efficiency} for details.  The repetitive nature of log files
gives them high compression ratios; the framework can read compressed log
files, simplifying the process of parsing them because they do not have to
be uncompressed to a temporary file before parsing begins.  When used
interactively, the framework also displays a progress bar to show how far
parsing has progressed through the log file and how long the remainder of
the parsing process is likely to take.  The progress bar is not as reliable
when parsing compressed log files, because a compressed block may
uncompress to a variable number of log lines; there is also variation
between the recognition and processing time of individual log lines, but
overall the progress bar is a useful addition.

@@

Actions are the component of this architecture responsible for processing
all of the inputs recognised by rules; in \parsername{} they reconstruct
the journey each mail takes through Postfix, dealing with all the
complications and difficulties that arise.  \parsername{} has
\numberOFactions{} actions and \numberOFrules{} rules, with an uneven
distribution of rules to actions as shown in \graphref{Distribution of
rules per action}.  Unsurprisingly, the action with the most associated
rules is \action{DELIVERY\_REJECTED}, the action that processes Postfix
rejecting a mail delivery attempt; next is \action{SAVE\_DATA}, the action
that saves useful information without doing any other processing.  The
third most common action is, perhaps surprisingly, \action{UNINTERESTING}:
this action does nothing when invoked, allowing uninteresting log lines to
be parsed without any effects (it does not imply that the input is
ungrammatical or unparsed).  Generally rules specifying the
\action{UNINTERESTING} action recognise log lines that are not associated
with a specific mail, e.g.\ notices about configuration files changing;
these log lines are recognised and processed so that the framework can warn
about unrecognised log lines, informing the user that they need to add to
or extend their ruleset.  Most of the remaining actions have only one or
two associated rules, because they will only ever have one or two log line
variants, e.g.\ all log lines showing that a remote client has connected
are recognised by a single rule and handled by the \action{CONNECT} action.

@@

    \item [DELIVERY\_REJECTED] Deals with Postfix rejecting an
        \acronym{SMTP} command from the remote client.  This is the action
        most frequently specified by rules, because so many different
        restrictions are used to reject delivery attempts.  This action is
        quite simple: if the log line contains a queueid, save the data
        captured by the rule's regex to the mail identified by that
        queueid; otherwise save it to the connection identified by the
        \acronym{pid} in the log line.

@@

    \item [MAIL\_DISCARDED] Sometimes mail is discarded by Postfix, e.g.\
        mail submitted locally that is larger than the limit configured by
        the administrator.  This action is used for those mails; it invokes
        the \action{COMMIT} action, but is a separate action to simplify
        further analysis.

@@

    \item [SMTPD\_DIED] Sometimes a \daemon{smtpd} dies, is killed by a
        signal, or exits unsuccessfully; the associated connection must be
        cleaned up, entered in the database if it has sufficient data, and
        deleted from the state tables.  In some cases the connection will
        not have enough data to satisfy the database schema, so it cannot
        be entered into the database for future analysis; unfortunately
        this means that the small amount of data that has been gathered
        will be lost.

@@

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible; one of the few limitations is that \parsername{} is
written in Perl, so new actions must also be written in Perl.  The
implementer writes a subroutine that accepts the standard arguments given
to actions, and registers it with the framework by calling the framework's
\texttt{add\_actions()} subroutine.  The new action must be registered
before the rules are loaded, because it is an error for a rule to specify
an unregistered action; this helps catch mistakes made when adding new
rules.  No other work is required from the implementer to integrate the
action into the parser; all of their attention and effort can be focused on
correctly implementing their new action.  The action may need to extend the
list of valid combinations described in \sectionref{out of order log lines}
if the new action creates a different set of acceptable programs, but this
would only be necessary if the new action processes log lines from Postfix
components \parsername{} currently does not have rules for, e.g.\
\daemon{virtual} or \daemon{lmtp}.

@@

There are other attributes that may be required to make a rule work
correctly: connection\_data, result\_data, priority, or restriction\_name.
In general it will only become clear that connection\_data or result\_data
are required when \parsername{} warns about an entry in the connections or
results tables that is missing some required fields, because values for the
missing fields are not present in any of the log lines.  For example, the
rule recognising the \daemon{pickup} component processing a mail sets the
client\_hostname to \texttt{localhost} and the client\_ip to
\texttt{127.0.0.1}, because the mail originates on the local server.  If
the new rule deliberately uses the architecture's overlapping rules feature
it will be obvious that the priority field needs to be set, on this rule
and possibly others; the priority field may be needed on unintentionally
overlapping rules too, but that is more difficult to determine.  Finally
the restriction\_name field should be set when the rule's action is
\action{DELIVERY\_REJECTED}; hopefully the name of the restriction will be
clear from the content of the log line.

@@

Rule conditions as part of the architecture have already been documented in
\sectionref{rule conditions in architecture}, and they can be very complex
and difficult to evaluate.  In contrast, \parsername{} uses very simple
rule conditions: each rule has a \texttt{program} attribute that specifies
the Postfix component whose log lines it recognises, and each log line
contains the name of the Postfix component that produced it; when trying to
recognise log lines, the framework will only use rules where the two are
equal.  This avoids needlessly trying rules that will not recognise the log
line, or worse, might recognise it unintentionally.  In addition the
framework supports generic rules, whose program attribute is ``*''; these
will be used if none of the program-specific rules recognise the log line.
If none of the rules are successful the framework will warn the user,
informing them that they need to augment their ruleset, and alerting them
that because \parsername{} failed to recognise some of their log lines, the
results stored in the database may be incomplete.

@@

    \item XXX FURTHER COMPLICATIONS MAY ARISE\@.

    \item XXX DATABASE GROWS WITHOUT BOUNDS\@.
